Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2017-12-26 survey/survey.tex
progress on intro
intro kind of done

\begin{comment}
\begin{quote}
  Ed Fredkin pursued the idea that information must be finite in
  density. One day, he announced that things must be even more simple
  than that. He said that he was going to assume that information
  itself is conserved. “You’re out of you mind, Ed.” I
  pronounced. “That’s completely ridiculous. Nothing could happen in
  such a world. There couldn’t even be logical gates. No decisions
  could ever be made.” But when Fredkin gets one of his ideas, he’s
  quite immune to objections like that; indeed, they fuel him with
  energy. Soon he went on to assume that information processing must
  also be reversible — and invented what’s now called the Fredkin
  gate. (Minsky 1999)

  In other terms, what is so good in logic that quantum physics should
  obey?  Can't we imagine that our conceptions about logic are wrong,
  so wrong that they are unable to cope with the quantum miracle?
  [\ldots] Instead of teaching logic to nature, it is more reasonable
  to learn from her. Instead of interpreting quantum into logic, we
  shall interpret logic into quantum (Girard 2007).
\end{quote}
\end{comment}

%%%%%%%%%
\paragraph*{Maxwell's Demon.}
To fully appreciate the missing principle of ``reversibility'' in
conventional computing, we go back to an old thought experiment by
J. C. Maxwell. The details are codified in a letter that Maxwell wrote
to P. G Tait in 1867 -- the letter, whose ideas are now known as
\emph{Maxwell's Demon}, tells of a thought experiment that seems to
indicate that intelligent beings can somehow violate the second law of
thermodynamics, thereby violating physics itself. Many resolutions
were offered for this conundrum (for a compilation, see the book by
Leff and Rex~\cite{leff1990}), but none withstood careful scrutiny
until the establishment of \emph{Landauer's Principle} in 1961 -- a
principle whose experimental validation happened recently in
2012~\cite{berut2012experimental}.

@@

%% (Cite Bennett and the various `thermodynamics of computation'  
%% papers here.)

@@

\amr{wavefront}

%%%%%%%%%
\paragraph*{Origins of Computational Models.}
Aspects of computation are
quite old. M\"{u}ller (1786) first conceived of the idea of a
``difference machine'', which Babbage (1819--1822) was able to
construct. There are other computer precursors as well -- the first
stored programs were actually for looms, most notably those of
Bouchon (1725) which operated on a paper tape, and Jacquard (1804)
which operated by chains of punched cards.

But it was Alan Turing's seminal work in 1936 which established the idea
that computation has a formal interpretation and that all
computability can be captured within a formal system. Implicit in this
achievement however is the idea that abstract models of computation
are just that -- \emph{abstractions of computation realized in the
physical world.}  In fact, one of the major achievements of Computer
Science has been the development of abstract models of computation
that shield the discipline from the underlying technology. As
effective as these models have been, one must note, however, that they
\emph{embody several implicit physical assumptions}.  As Tommaso
Toffoli explains in his influential 1980 paper:

{\begin{quote} Mathematical models of computation are abstract
  constructions, by their nature unfettered by physical laws. However,
  if these models are to give indications that are relevant to
  concrete computing, they must somehow capture, albeit in a selective
  and stylized way, certain general physical restrictions to which all
  concrete computing processes are subjected~\cite{toffoli:1980}.
\end{quote}}

Our logic of programs and our hardware are based on Boolean Logic
-- which really
ought to be called \emph{Piercean logic}, as Boole had a rather different
conception of logic which is rather far from our current models.
Nevertheless, going back
to Boole's 1853 book entitled \emph{An Investigation of the Laws of Thought,
  on which are Founded the Mathematical Theories of Logic and
  Probabilities}. The opening sentence of Ch.~1 is:
\begin{quote}
  The design of the following treatise is to investigate the fundamental laws
  of those operations of the mind by which reasoning is performed; \ldots
\end{quote}
A few chapters later, we find:
\begin{quote}
  \textbf{Proposition IV.}  That axiom of metaphysicians which is termed the
  principle of contradiction, and which affirms that it is impossible for any
  being to possess a quality, and at the same time not to possess it, is a
  consequence of the fundamental law of thought, whose expression is $x^2 =
  x$.
\end{quote}
A detailed historical analysis of Boole's ideas are beyond our scope.
The above quotes, however, should convey the idea that our notions of
computation generally date back to ideas that were thought reasonable
before \textasciitilde 1900.

It is conventional to base
the theory of computation and complexity on the Turing Machine.
Going back to Turing's 1936 article \emph{On Computable Numbers, with
  an Application to the Entscheidungsproblem,}, the opening sentence of
Sec. 1 is:
\begin{quote}
  We have said that the computable numbers are those whose decimals are
  calculable by finite means \ldots the justification lies in the fact that
  the human memory is necessarily limited.
\end{quote}
In Sec. 9, we find:
\begin{quote}
I think it is reasonable to suppose that they can only be squares
whose distance from the closest of the immediately previously observed
squares does not exceed a certain fixed amount.
\end{quote}
A detailed historical account is again beyond our scope. The quotes above
should convey the ideas that our theories
of computation and complexity are based on some assumptions that
Turing found reasonable in 1936. Though it is worth noting that these assumptions
are both physical (on distances) and metaphysical (on restrictions of the
mind).  If we take the human mind to be a physical ``machine'' which does
computation, then when both of the above assumptions are translated to the
language of physics, they embody what is known as the ``Bekenstein bound:''
which is an upper limit on the amount of information that can be contained
within a given finite region of space.

One can examine more and more cases but the general story should be
clear. Our abstractions were made up based on a certain understanding of the
laws of physics. In particular, our understanding of physics has evolved
tremendously since 1900!  Thus it is time to revisit these abstractions,
especially with respect to quantum mechanics.
In the words of Girard:
\begin{quote}
  In other terms, what is so good in logic that quantum physics should obey?
  Can't we imagine that our conceptions about logic are wrong, so wrong that
  they are unable to cope with the quantum miracle?
  \\
  Instead of teaching logic to nature, it is more reasonable to learn
  from her. Instead of interpreting quantum into logic, we shall
  interpret logic into quantum (Girard 2007).
\end{quote}

\begin{comment}
\jc{The reason I commented this out is that it is under-justified. The
reader will simply not understand what these next few lines are really
saying.}
Indeed one should take the physical principles underlying quantum mechanics,
the most successful physical theory known to us and adapt computation to
``learn'' from these principles. To illustrate the depth of our crisis, Scott
Aaronson, Umesh Vazirani, and others have proposed the following puzzle.

One of these wild claims must be true!:
\begin{itemize}
\item the extended Church-Turing thesis is false, or
\item quantum physics is false, or
\item there is an efficient classical algorithm for factoring
\end{itemize}
Indeed, if quantum physics is correct then there is an efficient quantum
algorithm for factoring (Shor). If there is no efficient classical algorithm
for factoring then the extended Church-Turing thesis is false.
\end{comment}

%%%%%%%%%
\paragraph*{From Classical to Quantum.}
Conventional classical models of computation, including boolean
circuits, the Turing machine, and the $\lambda$-calculus, are founded on
primitives which correspond to \emph{irreversible} physical processes which
gratuitously erase information.  For example, a \emph{nand} gate is an
irreversible logical operation in the sense that its inputs cannot generally
be recovered from observing its output, and so is the operation of overriding
a cell on a Turing machine tape with a new symbol, and so is a
$\beta$-reduction which typically erases or duplicates values in a way that
is destructive and irreversible.

These irreversible abstractions chosen as the basis of our
models of computing are at odds with the underlying
reversible physical reality. The information theoretic underpinnings
of physics is best understood through \emph{Landauer's
principle}~\cite{Landauer:1961}. As mentioned previously in the
introduction, Landauer, through his analysis of
Maxwell's Demon, established the idea that \emph{information is a
physical quantity} in much the sense that one would think of an
electron as a physical quantity. More precisely, his principle implies
that a certain minimum amount of
thermodynamic work has to be carried out to erase one bit of
information. Experimental verification of the Landauer principle
happened in recent years~\cite{berut2012experimental}. If information
has a physical significance, then it is subject to conservation laws
just as mass and energy are.

%%%%%%%%%
\paragraph*{Conservation of Information.}
To make this idea concrete and to provide a taste of the applications
it opens, consider a tiny 2-bit password = \verb|"10"|. The password
checker looks like:

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Programs as Reversible Deformations}

To better understand conservation of information, we can look
for analogous ideas outside of computing. Turning again to physics,
but this time at the macro scale, what does it mean to transform an
object in such a way that we do not lose the fundamental character
of the object?

For rigid objects (like a chair), the only such transformations are
translations and rotations. But what about something more flexible,
such as a water balloon?  They can be \textbf{deformed} in various ways,
but still retain their fundamental character -- as long as we do not
puncture them or over-stretch them. Ignoring material characteristics
(i.e. over-stretching), what is special about these deformations, as well
as for translations and rotations, is that they correspond to
continuous maps, with a continuous inverse. In fact, even more is
true: they are analytic maps, with analytic inverses. For our purpose,
the most important part is that such maps are infinitely differentiable.
In other words, not only is there an inverse to the deformation, but
its derivative is also invertible, and so on.

When we look around, we find many different words for related
concepts: isomorphism, equivalence, sameness, equality, interchangeability,
comparability, and correspondence, to name a few. Some of these
are informal concepts, while others have formal mathematical meaning.
More important, even amongst the formal concepts, there are differences
-- which is why there are so many of them! Thus we seek a concept
which is neither too strong nor too weak, that will express when
some structured information should be treated as ``the same''.

We will take ``the same'' as a fundamental principle. We want
to then derive what it means for data, programs, program
transformations, as well as proofs / deductions, to be ``the
same'' -- in an manner consistent with preservation of 
information.

This stands in stark contrast with most current approaches to
reversible computation, which start from current models of
computation involving irreversible operations and try to find
various ways to \emph{patch things up} so as to be reversible.

The reader will have surely noticed how we have studiously avoided
defining ``the same''. Because there are many such notions, we
also need to walk our way through them to find the one which is
``just right''. We can draw an analogy with topology: in topology,
all point sets can always be equipped with either the discrete or
the indiscrete topology, but both of these extremes are rarely
useful. We will develop our working notion of ``sameness''
as we go through the various components that make up a
programming language.

Starting from a physical perspective, whatever our notion of data
is, we will be interested in programs as representing transformations
of that data which are reversible. In other words, we want our
programs-as-transformations to ``play well'' with the inherent notion
of ``sameness'' that our data will carry. Thus we need to start by
looking at what structure our data has, and that will help us define
an appropriate notion of program. Of course, when programs themselves
are data, things do get more complicated.  In the following sections,
we will look at different natural classes of data, and explore the
corresponding notion of reversible programs.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Technical Overview}

@@

Several intriguing areas of exploration remain and the book concludes
with discussion of some of these, along with other possible
applications (see Ch.~\ref{ch:discussion}).  Of these, the field of
reversible and Quantum computing is closely related and we survey
several connections in depth.  The exact connection between $\Pi$ and
linear logic~\cite{Girard87tcs} is not fully understood though it is
clear that both systems capture related but different notions of
resource usage. A closely related area is the connection with duality
of computation \cite{Filinski:89, DBLP:conf/icfp/CurienH00,
  10.1109/LICS.2010.23,Wadler:2003} --- i.e., the quest for a unifying
framework between `computations', values, continuations and their
logical counterparts. The investigation of duality gives rise to
negative and fractional types which have an intuitive semantics and
are reminiscent of negative information flow, superposition and
entanglement from Quantum Physics~\cite{piee}.

>>>>

%%%%%%%%%
\paragraph*{Maxwell's Demon.}
To fully appreciate the missing principle of ``reversibility'' in
conventional computing, we go back to an old thought experiment by
J. C. Maxwell. The details are codified in a letter that Maxwell wrote
to P. G Tait in 1867 -- the letter, whose ideas are now known as
\emph{Maxwell's Demon}, tells of a thought experiment that seems to
indicate that intelligent beings can somehow violate the second law of
thermodynamics, thereby violating physics itself. Many resolutions
were offered for this conundrum (for a compilation, see the book by
Leff and Rex~\cite{leff1990}), but none withstood careful scrutiny
until the establishment of \emph{Landauer's Principle} in
1961~\cite{Landauer:1961} -- a principle whose experimental validation
happened recently in 2012~\cite{berut2012experimental}.

@@

@@

%%%%%%%%%
\paragraph*{Origins of Computational Models.}
Machines that ``compute'' are quite old. M\"{u}ller (1786) first
conceived of the idea of a ``difference machine'', which Babbage
(1819--1822) was able to construct. There are other computer
precursors as well -- the first stored programs were actually for
looms, most notably those of Bouchon (1725) which operated on a paper
tape, and Jacquard (1804) which operated by chains of punched cards.
Logical operations in high-level programming languages and in our
hardware are based on Boolean Logic -- which really ought to be called
\emph{Piercean logic}, as Boole had a rather different conception of
logic which is rather far from our current models.  Nevertheless,
going back to Boole's 1853 book entitled \emph{An Investigation of the
  Laws of Thought, on which are Founded the Mathematical Theories of
  Logic and Probabilities}. The opening sentence of Ch.~1 is:
\begin{quote}
  The design of the following treatise is to investigate the fundamental laws
  of those operations of the mind by which reasoning is performed; \ldots
\end{quote}
which clearly identifies the ``source'' of the logical laws as
mirroring Boole's understanding of human reasoning. A few chapters
later, we find:
\begin{quote}
  \textbf{Proposition IV.}  That axiom of metaphysicians which is termed the
  principle of contradiction, and which affirms that it is impossible for any
  being to possess a quality, and at the same time not to possess it, is a
  consequence of the fundamental law of thought, whose expression is $x^2 =
  x$.
\end{quote}
This ``law'' is reasonable in a classical world but is violated by
postulates of quantum mechanics. Although a detailed historical
analysis of Boole's ideas in the light of modern physics is beyond our
scope, the above quotes should convey the idea that our elementary
computing notions date back to ideas that were thought reasonable
before \textasciitilde 1900.

The 20th century witnessed the real advent of computer science as a
formal discipline. One of the pioneering works was Alan Turing's
seminal paper in 1936 which established the idea that computation has
a formal interpretation and that all computability can be captured
within a formal system. Implicit in this achievement however is the
idea that abstract models of computation are just that --
\emph{abstractions of computation realized in the physical world.}
Indeed, going back to Turing's 1936 article \emph{On Computable
  Numbers, with an Application to the Entscheidungsproblem,}, the
opening sentence of Sec. 1 is:
\begin{quote}
  We have said that the computable numbers are those whose decimals are
  calculable by finite means \ldots the justification lies in the fact that
  the human memory is necessarily limited.
\end{quote}
In Sec. 9, we find:
\begin{quote}
  I think it is reasonable to suppose that they can only be squares
  whose distance from the closest of the immediately previously
  observed squares does not exceed a certain fixed amount.
\end{quote}
It is worth noting that these assumptions are both physical (on
distances) and metaphysical (on restrictions of the mind).  If we take
the human mind to be a physical ``machine'' which does computation,
then when both of the above assumptions are translated to the language
of physics, they embody what is known as the ``Bekenstein bound:''
which is an upper limit on the amount of information that can be
contained within a given finite region of space. A detailed historical
account of these ideas in the context of modern physics is again
beyond our scope. However, the quotes above, like the ones before,
should convey the ideas that our theories of computation and
complexity are based on some assumptions that Turing and others found
reasonable in the 1930s.

To summarize, a major achievement of computer science has been the
development of abstract models of computation that shield the
discipline from rapid changes in the underlying technology. Yet, as
effective as these models have been, one must note that they
\emph{embody several implicit physical assumptions} and these
assumptions are based on a certain understanding of the laws of
physics.

Our understanding of physics has evolved tremendously since 1900!
Thus it is time to revisit these abstractions, especially with respect
to quantum mechanics.  Indeed one should take the physical principles
underlying quantum mechanics, the most successful physical theory
known to us and adapt computation to ``learn'' from these principles.
In the words of Girard:
\begin{quote}
  In other terms, what is so good in logic that quantum physics should
  obey?  Can't we imagine that our conceptions about logic are wrong,
  so wrong that they are unable to cope with the quantum miracle?
  [\ldots] Instead of teaching logic to nature, it is more reasonable
  to learn from her. Instead of interpreting quantum into logic, we
  shall interpret logic into quantum (Girard 2007).
\end{quote}

There are, in fact, many different principles of physics which are at
odds with our current models of computation. In this paper, we will
view data as an explicit representation of \emph{information},
programs as processes that transform information in a reversible way,
i.e., processes that are subject to the physical principle of
\emph{conservation of information.} We will formalize this idea and
follow its consequences, which will turn out to be far reaching.

%%%%%%%%%
\paragraph*{Programs as Reversible Deformations.}

To better understand the essence of ``conservation of information'' in
the context of computing, we first look for analogous ideas in
physics, but this time at the macro scale. Viewing information as a
physical object, what does it mean to transform an object in such a
way that we do not lose its fundamental character?

For rigid objects (like a chair), the only such transformations are
translations and rotations. But what about something more flexible,
with multiple representations, such as a water balloon?  Such objects
can be \emph{deformed} in various ways, but still retain their
fundamental character -- as long as we do not puncture them or
over-stretch them. Ignoring material characteristics
(i.e. over-stretching), what is special about these deformations, as
well as for translations and rotations, is that they correspond to
continuous maps, with a continuous inverse. In fact, even more is
true: they are analytic maps, with analytic inverses. For our purpose,
the most important part is that such maps are infinitely
differentiable.  In other words, not only is there an inverse to the
deformation, but its derivative is also invertible, and so on.

When we look around, we find many different words for related
concepts: isomorphism, equivalence, sameness, equality,
interchangeability, comparability, and correspondence, to name a
few. Some of these are informal concepts, while others have formal
mathematical meaning.  More important, even amongst the formal
concepts, there are differences -- which is why there are so many of
them! Because there are many such notions, we also need to walk our
way through them to find the one which is ``just right''. Thus we seek
a concept which is neither too strong nor too weak, that will express
when some structured information should be treated as ``the same''.
We can draw an analogy with topology: in topology, all point sets can
always be equipped with either the discrete or the indiscrete
topology, but both of these extremes are rarely useful. We will
develop our working notion of ``sameness'' as we go through the
various components that make up a programming language.

Starting from the physical perspective, whatever our notion of data
is, we will be interested in programs as representing transformations
of that data which are reversible. In other words, we want our
programs-as-transformations to ``play well'' with the inherent notion
of ``sameness'' that our data will carry. Thus we need to start by
looking at what structure our data has, and that will help us define
an appropriate notion of program. Of course, when programs themselves
are data, things do get more complicated.  In the following sections,
we will look at different natural classes of data, and explore the
corresponding notion of reversible programs.

To summarize, we will take ``the same'' as a fundamental principle and
derive what it means for data, programs, program transformations, as
well as proofs / deductions, to be ``the same'' -- in an manner
consistent with preservation of information. This stands in stark
contrast with most current approaches to reversible computation, which
start from current models of computation involving irreversible
operations and try to find various ways to \emph{patch things up} so
as to be reversible.

@@

Implement Fredkin gate and mention the following:

\begin{quote}
  Ed Fredkin pursued the idea that information must be finite in
  density. One day, he announced that things must be even more simple
  than that. He said that he was going to assume that information
  itself is conserved. “You’re out of you mind, Ed.” I
  pronounced. “That’s completely ridiculous. Nothing could happen in
  such a world. There couldn’t even be logical gates. No decisions
  could ever be made.” But when Fredkin gets one of his ideas, he’s
  quite immune to objections like that; indeed, they fuel him with
  energy. Soon he went on to assume that information processing must
  also be reversible — and invented what’s now called the Fredkin
  gate. (Minsky 1999)
\end{quote}

@@

\jc{The reason I commented this out is that it is under-justified. The  
reader will simply not understand what these next few lines are really
saying.}
Indeed one should take the physical principles underlying quantum mechanics,
the most successful physical theory known to us and adapt computation to
``learn'' from these principles. To illustrate the depth of our crisis, Scott
Aaronson, Umesh Vazirani, and others have proposed the following puzzle.

One of these wild claims must be true!:
\begin{itemize}
\item the extended Church-Turing thesis is false, or
\item quantum physics is false, or
\item there is an efficient classical algorithm for factoring
\end{itemize}
Indeed, if quantum physics is correct then there is an efficient quantum
algorithm for factoring (Shor). If there is no efficient classical algorithm
for factoring then the extended Church-Turing thesis is false.

@@

%%%%%%%%%
\subsection{Technical Overview}

@@

Several intriguing areas of exploration remain and the book concludes
with discussion of some of these, along with other possible
applications.  Of these, the field of reversible and Quantum computing
is closely related and we survey several connections in depth.  The
exact connection between $\Pi$ and linear logic~\cite{Girard87tcs} is
not fully understood though it is clear that both systems capture
related but different notions of resource usage. A closely related
area is the connection with duality of computation \cite{Filinski:89,
  DBLP:conf/icfp/CurienH00, 10.1109/LICS.2010.23,Wadler:2003} ---
i.e., the quest for a unifying framework between `computations',
values, continuations and their logical counterparts. The
investigation of duality gives rise to negative and fractional types
which have an intuitive semantics and are reminiscent of negative
information flow, superposition and entanglement from Quantum
Physics~\cite{piee}.

%%%%%%%%%
\paragraph*{Conservation of Information.}

If information has a physical significance, then it is subject to
conservation laws just as mass and energy are.  To make this idea
concrete and to provide a taste of the applications it opens, consider
a tiny 2-bit password = \verb|"10"|. The password checker looks like:

\textbf{punch line: allowable operations that computers can perform
  while embracing the laws of physics are operations that preserve
  everthing in abstract space of ``data''; doughnut into cup kind of
  transformations.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
